Associations between Chlorophyll *a* and various Microcystin-LR Health Advisory

### 2015-12-02 manuscript/manuscript.Rmd
closes #4 - Anne's comments.  Also added an addition table which needs to be created.

##Conditional Probability Analysis
We used a conditional probability analysis (CPA) approach to explore associations between chlorophyll *a* concentrations and World Health Organization (WHO) and USEPA microcystin-LR health advisory levels [@paul2011probability].  Many levels have been suggested (Table \ref{tab:microcystin_levels}), but lakes with higher microcystin-LR concentrations in the NLA were rare. Only `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=10])/nrow(nla_dat)),2)` % of lakes sampled had a concentration greater than 10.  Thus, for this analysis we focus on the microcystin concentrations that are better represented in the NLA data. These were  0.3, 1, 1.6, and 2 $\mu$g/L.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)`.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r round(median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)` and the mean was `r round(mean(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)`.  Of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA childrens drinking water standard, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA adult drinking water standard,`r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO drinking water standard,and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO recreational standard. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)` $\mu$g/L.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` and the mean was `r round(mean(nla_dat$CHLA,na.rm=T),2)`.  The associations between chlorophyll *a* and the upper confidence interval with a conditional probability of 50% were `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]` for 0.3, 1.0, 1.6 and 2.0 $\mu$g/L advisory levels, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).

@@

This is the case as the probability of exceeding each of the four tested health advisory levels increases as a function of chlorophyll *a* concentration (Figure \ref{fig:multi_cp_plot}).  We use this association to identify chlorophyll *a* concentrations that are associated with greater than even odds of exceeding a given health advisory level (Table \ref{tab:mc_chla_table}).  These represent `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[1]])/length(nla_dat$CHLA)),1)`%, `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[2]])/length(nla_dat$CHLA)),1)`%,
`r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[3]])/length(nla_dat$CHLA)),1)`%, and `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[4]])/length(nla_dat$CHLA)),1)`% of sample lakes for the U.S EPA childrens drinking water, the WHO drinking water, the USEPA Adult drinking water, and the WHO recreational advisories, respecitvely.

@@

There are numerous possible uses for the chlorophyll *a* and microcystin-LR advisory cut-off values.  First, in the absence of microcystin-LR measurements, exceedence of the chlorophyll *a* concentrations could be a trigger for further actions.  Given that there is uncertainity around these chlorophyll *a* cutoffs the best case scenario would be to monitor for chlorophyll and in the event of exceeding a target concentration take water samples and have those samples tested in a lab for microcystin-LR.

@@

```{r epa_child_cp_plot, echo=FALSE, fig.width=6, fig.cap="Conditional probability plots showing association between the probability of exceeding various microcystin-LR (MLR) health advisory Levels. A.) Plot for USEPA childrens drinking water advisory (0.3 $\\mu$g/L). B.) Plot for WHO drinking water advisory (1 $\\mu$g/L). C.) Plot for USEPA adult drinking water advisory (1.6 $\\mu$g/L). D.) Plot for WHO recreational advisory (2 $\\mu$g/L). \\label{fig:multi_cp_plot}"}
multiplot(
  plot_cp(epa_child_cp,x = "Log 10 Chlorophyll", 
          y = "Conditional Probability",
          title="A."),
  plot_cp(who_drink_cp, x = "Log 10 Chlorophyll", 
          y = "Conditional Probability", 
          title="B."),
  plot_cp(epa_adult_cp, x = "Log 10 Chlorophyll", 
          y = "Conditional Probability", 
          title="C."),
  plot_cp(who_rec_low1_cp, x = "Log 10 Chlorophyll", 
          y = "Conditional Probability", 
          title="D."),
  cols=2
)
```

@@

```{r mc_chla_table, results='asis', echo=FALSE, tab.cap="Chlorophyll \\textit{a} concentrations that are associated with a 50% probability of exceeding a microcystin-LR health advisory concentration. \\label{tab:mc_chla_table}"}
mc_chla_tbl <- data.frame(Source=mc_names,Type=mc_type,Microcystin=mc_guides,Chlorophyll=mc_chla)[1:4,]
#pandoc.table(mc_chla_tbl,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
knitr::kable(mc_chla_tbl)
```

>>>>

##Conditional Probability Analysis
We used a conditional probability analysis (CPA) approach to explore associations between chlorophyll *a* concentrations and World Health Organization (WHO) and USEPA microcystin-LR health advisory levels [@paul2011probability].  Many levels have been suggested (Table \ref{tab:microcystin_levels}), but lakes with higher microcystin-LR concentrations in the NLA were rare. Only `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=10])/nrow(nla_dat)),2)` % of lakes sampled had a concentration greater than 10.  Thus, for this analysis we focus on the microcystin concentrations that are better represented in the NLA data. These were the USEPA children's drinking water advisory level of  0.3 $\mu$g/L (USEPA Child), the WHO drinking water advisory level of 1 $\mu$g/L (WHO Drinking), the USEPA adult drinking water advisory level of 1.6 $\mu$g/L (USEPA Adult), and the WHO recreational, low probability of effect advisory level of 2 $\mu$g/L (WHO Recreational).

@@

In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)` $\mu$g/L.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r round(median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)` and the mean was `r round(mean(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)`.  Of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA Child level, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA Adult level,`r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO Drinking level,and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO Recreational level. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)` $\mu$g/L.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` $\mu$g/L and the mean was `r round(mean(nla_dat$CHLA,na.rm=T),2)` $\mu$g/L.  The associations between chlorophyll *a* and the upper confidence interval across a range of conditional probability values is shown in Table \ref{tab:mc_chla_table}.  Specific chlorophyll *a* that are associated with greater than even odds of exceeding the advisory levels were `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]` for 0.3, 1.0, 1.6 and 2.0 $\mu$g/L advisory levels, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).

@@

This is the case as the probability of exceeding each of the four tested health advisory levels increases as a function of chlorophyll *a* concentration (Figure \ref{fig:multi_cp_plot}).  We use this association to identify chlorophyll *a* concentrations that are associated with greater than even odds of exceeding a given health advisory level (Table \ref{tab:mc_chla_table}).  These represent `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[1]])/length(nla_dat$CHLA)),1)`%, `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[2]])/length(nla_dat$CHLA)),1)`%,
`r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[3]])/length(nla_dat$CHLA)),1)`%, and `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[4]])/length(nla_dat$CHLA)),1)`% of sample lakes for the USEPA Child, the WHO Drinking, the USEPA Adult, and the WHO recreational levels, respecitvely.

@@

There are numerous possible uses for the chlorophyll *a* and microcystin-LR advisory cut-off values.  First, in the absence of microcystin-LR measurements, exceedence of the chlorophyll *a* concentrations could be a trigger for further actions.  Given that there is uncertainity around these chlorophyll *a* cutoffs the best case scenario would be to monitor for chlorophyll *a* and in the event of exceeding a target concentration take water samples and have those samples tested in a lab for microcystin-LR.

@@

```{r epa_child_cp_plot, echo=FALSE, fig.width=6, fig.cap="Conditional probability plots showing association between the probability of exceeding various microcystin-LR (MLR) health advisory Levels. A.) Plot for USEPA Child (0.3 $\\mu$g/L). B.) Plot for WHO Drinking (1 $\\mu$g/L). C.) Plot for USEPA Adult (1.6 $\\mu$g/L). D.) Plot for WHO Recreational (2 $\\mu$g/L). \\label{fig:multi_cp_plot}"}
multiplot(
  plot_cp(epa_child_cp,x = "Log 10 Chlorophyll", 
          y = "Conditional Probability",
          title="A."),
  plot_cp(who_drink_cp, x = "Log 10 Chlorophyll", 
          y = "Conditional Probability", 
          title="B."),
  plot_cp(epa_adult_cp, x = "Log 10 Chlorophyll", 
          y = "Conditional Probability", 
          title="C."),
  plot_cp(who_rec_low1_cp, x = "Log 10 Chlorophyll", 
          y = "Conditional Probability", 
          title="D."),
  cols=2
)
```

@@

```{r mc_chla_table, results='asis', echo=FALSE, tab.cap="Chlorophyll \\textit{a} concentrations that are associated with a 50% probability of exceeding a microcystin-LR health advisory concentration. \\label{tab:mc_chla_table}"}
mc_chla_tbl <- data.frame(Source=mc_names,Type=mc_type,Microcystin=mc_guides,Chlorophyll=mc_chla)[1:4,]
#pandoc.table(mc_chla_tbl,justify = "left" ,split.table = 120, split.cells  = c(20,20,20))
knitr::kable(mc_chla_tbl)
####AS OF 12/2/15 - REDO THIS WITH MULTIPLE COND PROBS!
```
